---
title: "Health risk project"
output: "word_document"
date: "2026-04-27"


---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

---
title: "Health Risk Modeling — CDC Survey Data"
author: "Ashley & John"
date: "2026-04-07"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(haven)
brfss_raw <- read_xpt("path_to_my _one_drive_dataset")
```

```{r}
dim(brfss_raw)
head(brfss_raw)
```

#Lets explore the data

```{r}

# 0. Packages
############################
library(haven)
library(dplyr)
library(ggplot2)
library(glmnet)
library(nnet)
library(randomForest)
library(caret)

# Inspect
dim(brfss_raw)
head(brfss_raw)
str(brfss_raw)

############################
# 2. Basic Exploration
############################
# Example variable checks
table(brfss_raw$CVDINFR4, useNA = "ifany")
table(brfss_raw$CVDCRHD4, useNA = "ifany")
table(brfss_raw$PHYSHLTH, useNA = "ifany")

# Histogram of physical health days
ggplot(brfss_raw, aes(x = PHYSHLTH)) +
  geom_histogram(binwidth = 1, fill = "steelblue") +
  theme_minimal() +
  ggtitle("Distribution of Physical Health Days")
```

```{r}
# 3. Clean Special Codes

# Helper function for BRFSS missing codes
clean_brfss <- function(x) {
  x[x %in% c(7, 9, 77, 88, 99, 777, 888, 999)] <- NA
  return(x)
}

brfss <- brfss_raw

# Clean variables of interest
vars_to_clean <- c(
  "PHYSHLTH", "MENTHLTH", "GENHLTH", "EXERANY2",
  "MEDCOST1", "CHECKUP1", "CVDINFR4", "CVDCRHD4",
  "SEXVAR", "_BMI5"
)

for (v in vars_to_clean) {
  if (v %in% names(brfss)) {
    brfss[[v]] <- clean_brfss(brfss[[v]])
  }
}
```

```{r}

# 4. Create Targets

# 4A. Classification target: heart disease
# 1 if heart attack OR coronary heart disease, 0 if both explicitly "No"
brfss <- brfss %>%
  mutate(
    heart_disease = case_when(
      CVDINFR4 == 1 | CVDCRHD4 == 1 ~ 1,
      CVDINFR4 == 2 & CVDCRHD4 == 2 ~ 0,
      TRUE ~ NA_real_
    )
  )

table(brfss$heart_disease, useNA = "ifany")

# 4B. Regression target: poor physical health days
# Keep values 0 to 30 only
brfss <- brfss %>%
  mutate(
    poor_health_days = ifelse(PHYSHLTH >= 0 & PHYSHLTH <= 30, PHYSHLTH, NA)
  )

summary(brfss$poor_health_days)

```

```{r}

# 5. Non-response Indicators
# Professor feedback: non-response may carry signal

brfss <- brfss_raw %>%
  mutate(
    nonresp_PHYSHLTH = ifelse(PHYSHLTH %in% c(77, 88, 99), 1, 0),
    nonresp_MENTHLTH = ifelse(MENTHLTH %in% c(77, 88, 99), 1, 0),
    nonresp_GENHLTH  = ifelse(GENHLTH  %in% c(7, 9), 1, 0)
  ) %>%
  bind_cols(
    brfss %>% select(heart_disease, poor_health_days)
  )
```

```{r}

# 6. Select Candidate Variables
# Keep this manageable for project clarity
candidate_vars <- c(
  "heart_disease",
  "poor_health_days",
  "GENHLTH",
  "PHYSHLTH",
  "MENTHLTH",
  "PRIMINS2",
  "PERSDOC3",
  "MEDCOST1",
  "CHECKUP1",

  "EXERANY2",
  "SMOKE100",
  "SMOKDAY2",
  "ECIGNOW3",
  "USENOW3",

  "SEXVAR",
  "_AGEG5YR",
  "MARITAL",
  "EDUCA",
  "INCOME3",
  "EMPLOY1",
  "VETERAN3",
  "RENTHOM1",

  "_BMI5",

  "DIABETE4",
  "CVDSTRK3",
  "CHCKDNY2",
  "CHCCOPD3",
  "ADDEPEV3",
  "HAVARTH4",
  "ASTHMA3",

  "DEAF",
  "BLIND",
  "DECIDE",
  "DIFFWALK",
  "DIFFALON",

  "LASTDEN4",
  "HADSIGM4",
  "COLNCNCR",

  "nonresp_PHYSHLTH",
  "nonresp_MENTHLTH",
  "nonresp_GENHLTH"
)

candidate_vars <- candidate_vars[candidate_vars %in% names(brfss)]

project_df <- brfss %>%
  select(all_of(candidate_vars))
```

```{r}
############################
# 7. Classification Dataset
############################
class_df <- project_df %>%
  filter(!is.na(heart_disease)) %>%
  mutate(heart_disease = factor(heart_disease))

# Remove rows with missing predictors for simple baseline version
class_df_complete <- na.omit(class_df)

dim(class_df_complete)
table(class_df_complete$heart_disease)

# Plot class imbalance
ggplot(class_df_complete, aes(x = heart_disease)) +
  geom_bar(fill = "tomato") +
  theme_minimal() +
  ggtitle("Heart Disease Distribution")
```

```{r}
############################
# 8. Train / Test / Calibration Split
############################
set.seed(730)

idx <- sample(seq_len(nrow(class_df_complete)))
class_df_complete <- class_df_complete[idx, ]

n <- nrow(class_df_complete)
n_train <- floor(0.6 * n)
n_test  <- floor(0.2 * n)
n_cal   <- n - n_train - n_test

train_class <- class_df_complete[1:n_train, ]
test_class  <- class_df_complete[(n_train + 1):(n_train + n_test), ]
cal_class   <- class_df_complete[(n_train + n_test + 1):n, ]

dim(train_class)
dim(test_class)
dim(cal_class)
```

```{r}
############################
# 9. LASSO Variable Selection for Classification
############################
x_train <- model.matrix(heart_disease ~ ., train_class)[, -1]
y_train <- train_class$heart_disease

cv_lasso_class <- cv.glmnet(
  x_train,
  y_train,
  family = "binomial",
  alpha = 1
)

plot(cv_lasso_class)

lambda_1se_class <- cv_lasso_class$lambda.1se
lambda_1se_class

coef_class <- coef(cv_lasso_class, s = "lambda.1se")
coef_class

selected_class <- rownames(coef_class)[as.vector(coef_class != 0)]
selected_class <- setdiff(selected_class, "(Intercept)")
selected_class
```

```{r}
############################
# 10. Refit Logistic Regression
############################
if (length(selected_class) > 0) {
  formula_class <- as.formula(
    paste("heart_disease ~", paste(selected_class, collapse = " + "))
  )

  logit_model <- glm(
    formula_class,
    data = train_class,
    family = binomial()
  )

  summary(logit_model)

  # Odds ratios
  odds_ratios <- exp(coef(logit_model))
  odds_ratios

  # Confidence intervals for odds ratios
  # Uncomment if needed
  # exp(confint(logit_model))
}
```

```{r}
############################
# 11. Classification Performance
############################
test_probs <- predict(logit_model, newdata = test_class, type = "response")
test_pred  <- ifelse(test_probs >= 0.5, 1, 0)
test_pred  <- factor(test_pred, levels = c(0, 1))

confusionMatrix(test_pred, test_class$heart_disease, positive = "1")

############################
# 12. Conformal Classification
############################
# Score-based conformal classification for binary case
cal_probs <- predict(logit_model, newdata = cal_class, type = "response")

# Nonconformity scores:
# if true label = 1, score = 1 - p
# if true label = 0, score = p
cal_scores <- ifelse(
  cal_class$heart_disease == 1,
  1 - cal_probs,
  cal_probs
)

alpha <- 0.05
q_hat <- quantile(cal_scores, 1 - alpha, na.rm = TRUE)
q_hat

# Build prediction sets on test data
test_probs_conf <- predict(logit_model, newdata = test_class, type = "response")

prediction_sets <- lapply(test_probs_conf, function(p) {
  include_1 <- (1 - p) <= q_hat
  include_0 <- p <= q_hat

  out <- c()
  if (include_0) out <- c(out, 0)
  if (include_1) out <- c(out, 1)
  return(out)
})

# Coverage
true_test <- as.numeric(as.character(test_class$heart_disease))
coverage_class <- mean(
  sapply(seq_along(prediction_sets), function(i) true_test[i] %in% prediction_sets[[i]])
)
coverage_class

# Average set size
avg_set_size_class <- mean(sapply(prediction_sets, length))
avg_set_size_class
```

```{r}
############################
# 13. Regression Dataset
############################
reg_df <- project_df %>%
  select(-heart_disease, -PHYSHLTH, -nonresp_PHYSHLTH) %>%
  filter(!is.na(poor_health_days)) %>%
  na.omit()

dim(reg_df)
summary(reg_df$poor_health_days)

ggplot(reg_df, aes(x = poor_health_days)) +
  geom_histogram(binwidth = 1, fill = "darkgreen") +
  theme_minimal() +
  ggtitle("Distribution of Poor Health Days")
```

```{r}
############################
# 14. Train / Test / Calibration Split for Regression
############################
set.seed(730)

idx_reg <- sample(seq_len(nrow(reg_df)))
reg_df <- reg_df[idx_reg, ]

n_reg <- nrow(reg_df)
n_train_reg <- floor(0.6 * n_reg)
n_test_reg  <- floor(0.2 * n_reg)

train_reg <- reg_df[1:n_train_reg, ]
test_reg  <- reg_df[(n_train_reg + 1):(n_train_reg + n_test_reg), ]
cal_reg   <- reg_df[(n_train_reg + n_test_reg + 1):n_reg, ]
```

```{r}
############################
# 15. LASSO for Regression
############################
x_train_reg <- model.matrix(poor_health_days ~ ., train_reg)[, -1]
y_train_reg <- train_reg$poor_health_days

cv_lasso_reg <- cv.glmnet(
  x_train_reg,
  y_train_reg,
  family = "gaussian",
  alpha = 1
)

plot(cv_lasso_reg)

lambda_1se_reg <- cv_lasso_reg$lambda.1se
lambda_1se_reg

coef_reg <- coef(cv_lasso_reg, s = "lambda.1se")
coef_reg

selected_reg <- rownames(coef_reg)[as.vector(coef_reg != 0)]
selected_reg <- setdiff(selected_reg, "(Intercept)")
selected_reg

############################
# 16. Refit Linear Regression
############################
if (length(selected_reg) > 0) {
  formula_reg <- as.formula(
    paste("poor_health_days ~", paste(selected_reg, collapse = " + "))
  )

  lm_model <- lm(formula_reg, data = train_reg)
  summary(lm_model)
}
```

```{r}
############################
# 17. Conformal Regression
############################
# Absolute residual conformal regression
cal_pred_reg <- predict(lm_model, newdata = cal_reg)
cal_scores_reg <- abs(cal_reg$poor_health_days - cal_pred_reg)

q_hat_reg <- quantile(cal_scores_reg, 1 - alpha, na.rm = TRUE)
q_hat_reg

test_pred_reg <- predict(lm_model, newdata = test_reg)

lower <- test_pred_reg - q_hat_reg
upper <- test_pred_reg + q_hat_reg

# Coverage
coverage_reg <- mean(
  test_reg$poor_health_days >= lower & test_reg$poor_health_days <= upper
)
coverage_reg

# Average interval width
avg_width_reg <- mean(upper - lower)
avg_width_reg
```

```{r}

# 19. Simple Summary Outputs
cat("\n===== CLASSIFICATION RESULTS =====\n")
cat("Selected variables (LASSO):\n")
print(selected_class)
cat("\nConformal classification coverage:", coverage_class, "\n")
cat("Average prediction set size:", avg_set_size_class, "\n")

cat("\n===== REGRESSION RESULTS =====\n")
cat("Selected variables (LASSO):\n")
print(selected_reg)
cat("\nConformal regression coverage:", coverage_reg, "\n")
cat("Average interval width:", avg_width_reg, "\n")
